---
title: "Data Quality Assessment"
author: "Sujit Sivadanam"
date: "11/30/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data Quality Function

```{r}
# Column completeness
col_complete <- function(dataframe){
  if(!is.data.frame(dataframe)){
    stop("the object is not a dataframe")
  }
  
  missing_count <- 0
  for(j in 1:ncol(dataframe)){
    for(i in 1:nrow(dataframe)){
      if(is.na(dataframe[i, j])){missing_count <- missing_count+1}
    }
  }
  complete_ratio <- 1 - missing_count/(nrow(dataframe)*ncol(dataframe))
  return(complete_ratio)
}
```

## Assessing Data Quality for Temp Data

```{r}
library(visdat)
library(naniar)
library(tidyverse)

# Analyzing for completeness
col_complete(temp.data.converted$airsv6)
col_complete(temp.data.converted$airsv7)
col_complete(temp.data.converted$ghcnv4)

# Analyzing for consistency
vis_miss(temp.data.converted$airsv6)
vis_miss(temp.data.converted$airsv7)
vis_miss(temp.data.converted$ghcnv4)
```

## Assessing Data Quality for Gas Data

```{r}
# Analyzing for completeness
col_complete(as.data.frame(gas.data.converted$ch4))
col_complete(as.data.frame(gas.data.converted$n2o))
col_complete(as.data.frame(gas.data.converted$sf6))

# Analyzing for consistency
vis_miss(as.data.frame(gas.data.converted$ch4))
vis_miss(as.data.frame(gas.data.converted$n2o))
vis_miss(as.data.frame(gas.data.converted$sf6))
```

## Outliers for Temp Data

```{r}
airsv6 <- subset(temp.data.converted$airsv6, select = -Year)
boxplot(subset(airsv6, select = c("Jan", "Feb", "Mar", "Apr")))
boxplot(subset(airsv6, select = c("May", "Jun", "Jul", "Aug")))
boxplot(subset(airsv6, select = c("Sep", "Oct", "Nov", "Dec")))
boxplot(subset(airsv6, select = c("J.D", "D.N", "DJF", "MAM", "JJA", "SON")))

airsv7 <- subset(temp.data.converted$airsv7, select = -Year)
boxplot(subset(airsv7, select = c("Jan", "Feb", "Mar", "Apr")))
boxplot(subset(airsv7, select = c("May", "Jun", "Jul", "Aug")))
boxplot(subset(airsv7, select = c("Sep", "Oct", "Nov", "Dec")))
boxplot(subset(airsv7, select = c("J.D", "D.N", "DJF", "MAM", "JJA", "SON")))

ghcnv4 <- subset(temp.data.converted$ghcnv4, select = -Year)
boxplot(subset(ghcnv4, select = c("Jan", "Feb", "Mar", "Apr")))
boxplot(subset(ghcnv4, select = c("May", "Jun", "Jul", "Aug")))
boxplot(subset(ghcnv4, select = c("Sep", "Oct", "Nov", "Dec")))
boxplot(subset(ghcnv4, select = c("J.D", "D.N", "DJF", "MAM", "JJA", "SON")))
```

## Outliers for Gas Data

```{r}
ch4 <- subset(as.data.frame(gas.data.converted$ch4), select = -Year)
n2o <- subset(as.data.frame(gas.data.converted$n2o), select = -Year)
sf6 <- subset(as.data.frame(gas.data.converted$sf6), select = -Year)

boxplot(subset(ch4, select = Uncertainty_avg))
boxplot(subset(ch4, select = Uncertainty_trend))

boxplot(subset(n2o, select = Uncertainty_avg))
boxplot(subset(n2o, select = Uncertainty_trend))

boxplot(subset(sf6, select = Uncertainty_avg))
boxplot(subset(sf6, select = Uncertainty_trend))
```

## Report

The ratios of column completeness for both the "airsv6" and "airsv7" datasets are both identically the same (0.9573935) while the ratio of column completeness for the "ghcnv4" dataset was just a bit higher (0.9899749) than the other two. In other words, the datasets are almost complete in that the few missing values are not too big of a concern. All three of the temp datsets are pretty consistent in terms of precision.

In the "airsv6" dataset, the months of January, May, June, & November each have one outlier. In the "airsv7" dataset, January, March, & June each have one outlier while November has five outliers.

For the gas data, the ratios of column completeness for all three data sets (ch4, n2o, sf6) are each 0.9927052, which is very good because these few missing values won't have a significant effect on the data analysis. Furthermore, each column of each of the three datasets are consistently precise. In terms of outliers, the "ch4" dataset has the most outliers.
